---
title: "Phyloseq_test"
author: "Simon Rayner"
date: '2023-08-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Phyloseq can handle the qza files output from Qiime2 and they claim to produce superior visualisations. 
To begin with, let's look at the sample dataset they use to ensure we get the same results.

I started trying to follow this tutorial, but using Bai's Qiime2 files from the 24 samples (8 nasal aspirate/swabs/saliva)

https://rdrr.io/github/jbisanz/qiime2R/f/README.md


```{r loadLibraries}
library(GUniFrac)
library(vegan)

library(devtools)
# LDM Requires devtools and "BiocParallel
# then:
#   install.packages("/Users/simonray/NetBeansProjects/LDM/LDM_6.0.tar.gz", repos=NULL)
library(LDM)
#BiocManager::install("phyloseq")
library(phyloseq)
library(tidyverse)
# devtools::install_github("jbisanz/qiime2R")
library(qiime2R)
#library(ggplot2)
library(reshape2)

# Rsamtools
```

## Data Import using qiime2R

### set path
```{r importWithQiime2R_setpath}
qiime2DataFolder<-"/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples"
```

### load metadata from Qiime2 file
```{r importWithQiime2R_metadata}
metadata<-read_q2metadata(file.path(qiime2DataFolder, "16s_paper_analysis__sample-metadata4R.tsv"))
```

### load sequence variants
```{r importWithQiime2R_seqvariants}
SVTable<-read_qza(file.path(qiime2DataFolder, "3-table.qza"))

```
This is appears to load correctly because
```
> names(SVTable)
```
gives us
```
 [1] "uuid"       "type"       "format"     "contents"   "version"    "data"       "provenance"
```
which is consistent with the name in the example here --> https://rdrr.io/github/jbisanz/qiime2R/f/README.md

Check some more

```{r importWithQiime2R_seqvariants_view}
SVTable$data[1:5,1:5] #show first 5 samples and first 5 taxa
```

## Load Taxonomy
```{r importWithQiime2R_taxonomy}
taxonomy<-read_qza(file.path(qiime2DataFolder,"4-taxonomy.qza"))
head(taxonomy$data)
```

## Load Tree
#The following throws an error because it is a Phyloseq function, not qiime2R and it expects a Newark tree.
```{r importWithQiime2R_tree}
rootedTree<-read_tree(file.path(qiime2DataFolder,"6-rooted-tree.qza"))
```

but, we can specify the tree if we use the `qza_to_phyloseq`function. 

```
    cnut<-qza_to_phyloseq(features="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/3-table.qza",
      tree="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/6-rooted-tree.qza",
      taxonomy="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/4-taxonomy.qza",
      metadata="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/16s_paper_analysis__sample-metadata4R.tsv")
```
But, can do the following?

```
    cnut<-qza_to_phyloseq(features=SVTable,
      tree="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/6-rooted-tree.qza",
      taxonomy=taxonomy,
      metadata=metadata)
```
No, i can't. I get the following error
```
Error in file.exists(file) : invalid 'file' argument
```

So, `qza_to_phyloseq` requires user to specify file paths. 

Go ahead and create a phyloseq object using `qza_to_phyloseq`

```{r createPhyloseqObjectFromFiles}
    physeqObject<-qza_to_phyloseq(features=file.path(qiime2DataFolder, "3-table.qza"),
      tree=file.path(qiime2DataFolder, "6-rooted-tree.qza"),
      taxonomy=file.path(qiime2DataFolder,"4-taxonomy.qza"),
      metadata=file.path(qiime2DataFolder, "16s_paper_analysis__sample-metadata4R.tsv"))
```


## make a bar plot of the Shannon Diversity by sample

`geom_point`
`theme_q2r()` 
`scale_shape_manual`
`scale_size_continuous`
`scale_color_discrete`

First, need to load the calculated Shannon values

```{r Shannon}
shannon<-read_qza(file.path(qiime2DataFolder,"/6-diversity-metrics/shannon_vector.qza")) 
# make the row names as a column so we can join with the phyloseq object
shannon<-shannon$data %>% rownames_to_column("SampleID") 


gplots::venn(list(metadata=metadata$SampleID, shannon=shannon$SampleID))
```



```{r ShannonPlot}
metadata2<-metadata %>% left_join(shannon)
metadata2$ss<-factor(metadata2$samplesource)
ggplot(metadata2, aes(x=ss, y=shannon_entropy)) + geom_boxplot(notch=FALSE) + stat_summary(fun.y=mean, geom="point", shape=23, size=4)  + geom_jitter(shape=16, position=position_jitter(0.2))
metadata2 %>% 
  ggplot(metadata2, aes(x=ss, y=shannon_entropy)) + geom_boxplot(notch=FALSE) + stat_summary(fun.y=mean, geom="point", shape=23, size=4)  + geom_jitter(shape=16, position=position_jitter(0.2))
metadata2 %>%  
  ggplot(aes(x="samplesource", y=shannon_entropy, fill='ss')) +
  #stat_summary(geom="bar", fun.data=mean_se, color="black") + #here black is the outline for the bars
  geom_jitter(shape=21, width=0.2, height=0) +
  coord_cartesian(ylim=c(2,7)) + # adjust y-axis
  #facet_grid(~'source') + # create a panel for each body site
  xlab("ss") +
  ylab("Shannon Diversity") +
  geom_boxplot() +
  #theme_q2r() +
  scale_fill_manual(values=c("cornflowerblue","indianred", "hotpink4", "magenta4")) + #specify custom colors
  theme(legend.position="none") #remove the legend as it isn't needed
  ggsave(file.path(qiime2DataFolder, "6-diversity-metrics/Shannon_by_abx.pdf"), height=3, width=4, device="pdf") # save a PDF 3 inches by 4 inches
```






Now try to create a phyloseq object from these

```{r importWithQiime2R4}
otuR<-otu_table(SVTable$data, taxa_are_rows = TRUE)
physeq2 = phyloseq(otuR, taxonomy)
plot_bar(physeq2, fill = "Phylum")

uwunifrac<-read_qza(file.path(qiime2DataFolder, "6-diversity-metrics/unweighted_unifrac_pcoa_results.qza"))
shannon<-read_qza(file.path(qiime2DataFolder, "6-diversity-metrics/shannon_vector.qza"))$data %>% rownames_to_column("SampleID") 
```

this imports from a saved dataframe for a sample dataset, but how to load qza

```{r sampleDataset, echo=FALSE}

data ("GlobalPatterns")

# This creates a Phyloseq object
GP <- prune_taxa(taxa_sums(GlobalPatterns) > 0, GlobalPatterns)

human <- get_variable(GP, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue")

sample_data(GP)$human <- factor(human)

sample_data(GP)$human<-ifelse(human==TRUE, "Human", "Not-Human")

#create and label human variable
head(sample_data(GP))

```

## Loading Qiime2 data


```{r loadQiime2Data}
# this loads the data into a list
SVTable<-read_qza("/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/3-table.qza")
otuR<-otu_table(SVTable$data, taxa_are_rows = TRUE)
# This is a large list, not a Phyloseq object

# for loading in R, need to add comment line for line 2 to describe data types for each column
# For example
# 
# sample-id	source
# #q2:types	Categorical.              <-- R requires this line (must start with #q2:types)
# 10-30-3A_S20	nasal_aspirate
# 
metadata<-read_q2metadata("/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/16s_paper_analysis__sample-metadata4R.tsv")
rootedTree<-read_tree("/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/6-rooted-tree.qza")
taxonomy<-read_qza("/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/4-taxonomy.qza")


SVTable<-read_qza("/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/3-table.qza")
otuR<-otu_table(SVTable$data, taxa_are_rows = TRUE)
physeq2 = phyloseq(otuR, taxonomy)
plot_bar(physeq2, fill = "Phylum")
```
Then build a Phyloseq object using the wrapper function `qza_to_phyloseq()`

```{r qiime2Load}

    cnut<-qza_to_phyloseq(features="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/3-table.qza", tree="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/6-rooted-tree.qza", taxonomy="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/4-taxonomy.qza", metadata="/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/16s_paper_analysis__sample-metadata4R.tsv")
```




```{r plotCrap}
uwunifrac$data$Vectors %>%
  select(SampleID, PC1, PC2) %>%
  left_join(metadata) %>%
  left_join(shannon) %>%
  #ggplot(aes(x=PC1, y=PC2, color=`source`, shape=`reported-antibiotic-usage`, size=shannon)) +
  ggplot(aes(x=PC1, y=PC2, color='source', size=shannon_entropy)) +
  geom_point(alpha=0.5) + #alpha controls transparency and helps when points are overlapping
  theme_q2r() +
  scale_shape_manual(values=c(16,1), name="source") + 
  #see http://www.sthda.com/sthda/RDoc/figure/graphs/r-plot-pch-symbols-points-in-r.png 
  # for numeric shape codes
  scale_size_continuous(name="Shannon Diversity") +
  scale_color_discrete(name="source")
  ggsave(file.path(qiime2DataFolder,"PCoA.pdf"), height=4, width=5, device="pdf") # save a PDF 3 inches by 4 inches
```


## Phyloseq

```{r pressure, echo=FALSE}
theme_set(theme_classic())

alpha_meas = c("Shannon")

plotGP <- plot_richness(GP, "human", "SampleType", measures=alpha_meas)
plotGP + geom_boxplot(data=plotGP$data, aes(human,value,color=NULL), alpha=0.1)+coord_cartesian(ylim = c(0, 7))
```



## AlphaDiversity

```{r Shannon}
shannon<-read_qza("/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples/6-diversity-metrics/shannon_vector.qza") 
shannon<-shannon$data %>% rownames_to_column("SampleID") 
# this moves the sample names to a new column that matches the metadata and allows them to be merged

gplots::venn(list(metadata=metadata$SampleID, shannon=shannon$SampleID))
```

```{r PCofA}
library(tidyverse)
library(qiime2R)

qiime2DataFolder<-"/Users/simonray/DropboxUiO/dropData/ngsdata/run1_24_saliva_nasal_samples"


metadata<-read_q2metadata(file.path(qiime2DataFolder, "16s_paper_analysis__sample-metadata4R.tsv"))
uwunifrac<-read_qza(file.path(qiime2DataFolder, "6-diversity-metrics/unweighted_unifrac_pcoa_results.qza"))
shannon<-read_qza(file.path(qiime2DataFolder, "6-diversity-metrics/shannon_vector.qza"))$data %>% rownames_to_column("SampleID") 

uwunifrac$data$Vectors %>%
  select(SampleID, PC1, PC2) %>%
  left_join(metadata) %>%
  left_join(shannon) %>%
  #ggplot(aes(x=PC1, y=PC2, color=`source`, shape=`reported-antibiotic-usage`, size=shannon)) +
  ggplot(aes(x=PC1, y=PC2, color='source', size=shannon_entropy)) +
  geom_point(alpha=0.5) + #alpha controls transparency and helps when points are overlapping
  theme_q2r() +
  scale_shape_manual(values=c(16,1), name="source") + 
  #see http://www.sthda.com/sthda/RDoc/figure/graphs/r-plot-pch-symbols-points-in-r.png 
  # for numeric shape codes
  scale_size_continuous(name="Shannon Diversity") +
  scale_color_discrete(name="source")
  ggsave(file.path(qiime2DataFolder,"PCoA.pdf"), height=4, width=5, device="pdf") # save a PDF 3 inches by 4 inches
```


Don't have enough variables to generate this plot (no longitudinal informatiom)
```{r ShannonPlot}
metadata2<-metadata %>% left_join(shannon)
metadata2$ss<-factor(metadata2$samplesource)
metadata2 %>%  
  ggplot(aes(x="ss", y=shannon_entropy, fill='ss')) +
  stat_summary(geom="bar", fun.data=mean_se, color="black") + #here black is the outline for the bars
  geom_jitter(shape=21, width=0.2, height=0) +
  coord_cartesian(ylim=c(2,7)) + # adjust y-axis
  #facet_grid(~'source') + # create a panel for each body site
  xlab("ss") +
  ylab("Shannon Diversity") +
  theme_q2r() +
  scale_fill_manual(values=c("cornflowerblue","indianred", "hotpink4", "magenta4")) + #specify custom colors
  theme(legend.position="none") #remove the legend as it isn't needed
  ggsave(file.path(qiime2DataFolder, "6-diversity-metrics/Shannon_by_abx.pdf"), height=3, width=4, device="pdf") # save a PDF 3 inches by 4 inches
```




```{r f}
theme_set(theme_classic())
alpha_meas = c("Shannon")

plotGP <- plot_richness(metadata2, "ss", "ss", measures=alpha_meas)

plot_bar(cnut, fill="Phylum")
plot_bar(cnut, fill="Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")
cnut@sam_data$samplesource
cnut_fraction <- merge_samples(cnut, "samplesource")
  plot_bar(cnut_fraction, fill = "Phylum") + 
  geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")

```